Stagnating Public University Finances and Faculty Composition

### 2022-08-04 text/paper.tex
rough first draft.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
    Cornell University, Department of Economics.}
    \thanks{I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, and Francine Blau for various comments on the project's direction,
    as well as seminar participants at Cornell University (Spring 2022) for helpful
    discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/finances-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.}}
% Title
\title{Stagnating Public University Finances and Faculty Composition}
\date{\today}
\rhead{\today}
\lhead{Stagnating Public University Finances and Faculty Composition}
% Begin
\begin{document}
\clearpage \maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
\noindent
\textbf{This needs re-doing}

When a worker's employer is exposed to demand shocks we may expect the resulting effect on the worker to depend on multiple economic factors, including size of the shock and the worker's outside options.
This paper studies the topic for a unique labour market: that of public university professors in the US.
State spending for public universities has secularly declined the last three decades, yet the isolated effects on professors are not yet clear.
Using an instrumental variables strategy to isolate shocks in public university revenues, I show that universities respond to negative revenue shocks by restricting employment among tenure-track and tenured professors -- possibly due to hiring and retirement channels, respectively.
Results using individual level data regarding every public university professor in Illinois are to follow.

\vspace{0.75cm}
\noindent\textbf{Keywords:} Public Sector Labor Markets, Higher Education, State and Local Budget and Expenditures

\vspace{0.5cm}
\noindent\textbf{JEL Codes:} J45, I23, H72
\end{abstract}
\newpage
\setcounter{page}{1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Starting the paper
\newpage
\doublespacing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Plan
\subsection{Plan}
\begin{enumerate}
    \item Document empirical trend about number of professors per students, possibly diverging between private and public unis
    \item Document trend of falling revenues per student in the public sector -- while the private sector is fine
    \item Pull these facts together with shift-share IV approach to attribute the trend to falling state support for PUs.
    \item Document the effect on individuals in Illinois, including local projections for the dynamic effects.
    \item Theory discussion on the goals of a public university: public university charters require a minimum provision of teaching that make teaching a constraint, and other activities (including research).
    Implies exmploying fewer instructors per student, or substitution towards cheaper instructors who focus exclusively on teaching. 
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
\section{Introduction}
\noindent
Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) the last few decades.
This decline has been shown to lead to worse education and later-life outcomes for students \citep{NBERw23736,NBERw27885}, yet it is not clear what institutional effects or mechanisms driving effects on students.
I show that four-year, degree-granting public universities (PUs) are exposed to tightening financial constraints thanks to falling state financial support (per student), and show this effects the composition of professors at PUs.
Falls in PU revenues, instrumented by state-level finance shocks, lead to a fall in the number of professors per student within a PU, driven by a fall in assistant or tenured professor counts.
US private universities are not exposed to similar financial constraints during this same time period, and do not exhibit any similar changes in the faculty composition, so that stagnating financial support for PUs have implications for the structure of higher education instruction and research in the US.

Universities in the US are widely considered the highest performing in the world,yet there are consequential differences between its universities that operate in the private sector and those established by state governments.
PUs are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average PU received around \$11,600 per enrolled student in 1990, yet only \$8,300 per enrolled student in 2020.\footnote{
    2020 figures are relatively similar to those for year preceding, despite the Covid-19 pnademic and associated financial shocks, since 2020 refers to the academic year 2019-2020 and so the yearly budget was decided in 2019.
    This paper does not directly analyse budget shocks related to the Covid-19 pandemic, or any other events in 2020 and following.
}
This fall is driven by stagnating funding provided by state governments, while total enrollment among PUs rose by 46\% over the same time period.
At the same time, the number of professors per student at private universities stayed relatively stable at 0.045, yet fell from around 0.04 to around 0.35 professors per student at PUs, driven by a fall of over 20\% in the ratio of associate and full (i.e. tenured) professors per student at PUs.\footnote{
    \autoref{fig:fte-perprof} and \autoref{sec:trends} document these trends between private and public universities in full detail.
}

I use a variant of the \cite{NBERw23736,NBERw27885} shift-share instrument for PU revenues from shocks to state appropriations to identify the effect on the faculty composition.
State appropriations for public universities are plausibly endogenous to many outcomes: state governments decide on yearly budgets for their higher education sector, and this process can be influenced by local financial conditions, or even public (and political representatives') perceptions of the state's higher education system.
The shift-share instrument interacts reliance on state funding in a base period with the yearly total appropriations per student in the entire state;
the instrument for yearly state appropriations addresses the endogeneity concerns if state government funding decisions are not associated with any outcomes for any one university in the state \citep{NBERw27885}.
\autoref{sec:approp-shocks} provides further evidence in favour of the exogeneity assumption of the instrument.

Briefly describe results.

Then write about the Illinois data as individual-level data, and empirical approach when it is fully specified.

@@

University professors the core component of this country's educational and research/innovation university system, yet their composition within universities and its relationship with higher education finances has so far not been studied.
\cite{brown2014endowment} presents the closest example by studying how university endowments react to negative financial shocks, exhibiting an association between negative endowment shocks and a fall in tenure-system faculty employed at the university.
\cite{abe2015implications} present a model for university administrator decision-making, and posit that a large component of variation in faculty salaries arises from university politics, and not only economic factors such as outside.
\cite{johnson2009jep,NBERc13879} both empirically document allocation of faculty between research and instruction, and \cite{hemelt2021math} document differences in instructional cost (per student) between departments, primarily thanks to department differences in class size and faculty salaries.
\cite{turner2014impact} document how universities reacted to the Great Recession of 2008, a severe financial shock to state finances, including implementing hiring freezes.
Indeed, Cornell University\footnote{
    Cornell University is, indeed, only partially public, so that this information is not free information, and cannot be presented with aggregate or verified sources of public data on such salaries.
} implemented both hiring freezes and even a nominal salary reduction for professors, in anticipation of oncoming financial shock in early 2020.\footnote{
    The salary cut was not permanent, as the oncoming financial shock turned out to not be as serious as projected, and the salary cut was returned to professors in the year 2021.
}

@@

\textbf{Add in an intro conclusion paragraph.}
``Furthermore, because schools that serve students from lower-income backgrounds are most affected by state appropriations cuts, reductions in state support have helped exacerbate inequality and stratification of outcomes in the postsecondary sector.'' \citep{NBERw27885}.

@@

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    This statement means that IPEDS does not necessarily cover the universe of US higher education institutions, yet thin practice every PU and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2020,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e. 1987 refers to the academic year that ran August 1986 to July 1987.
}
and provide information on the total revenues a university received from every source (including state governments), enrollment number, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990-2017, and manually combine with raw NCES data regarding the year 2018-2020 for all relevant variables.
    Figures for enrollment, faculty counts and salaries come from the raw NCES version of IPEDS for all years to address inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to four-year institutions, as these institutions adhere to a standardised concept of faculty profile -- that is, tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised in four-year, degree-granting institutions.
Additionally, for-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enroll at a four-year institution \citep{mountjoy2022}.\footnote{
    Furthermore, \cite{mountjoy2022} documents the effects of expanding two-year higher education access in the US separately for students who would have and those who would not have otherwise have attended college.
    The high rate of eventual enrollment at four-year institutions among two-year students explains the negative impact of increased access to two-year institutions on students who are diverted away from four-year institutions.
}

@@

\begin{table}[h!]
    \onehalfspacing
    \centering
    \caption{IPEDS Summary Statistics, PUs Panel 1987--2020}
    \makebox[\textwidth][c]{\input{tables/ipeds-summary-fte.tex}}
    \label{tab:ipeds-summary}
\end{table}

IPEDS provides information at the university level, yet lacks information on the distribution of salaries and professor count within a university.\footnote{
    IPEDS provides total paid to salary and total faculty employed per university-year, so that I can investigate average professor salary per university-year with IPEDS data.
    Unfortunately, it seems that this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period \citep{aau2021survey}.
}
To investigate the distribution, I integregate individual-level data for every PU professor in the state of Illinois between the years 2010-2020.
IBHED host the information;
Public Act 96-0266 (effective 1 January 2010) requires that each PU report to the IBHE the base salary and benefits of the president and all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the eight Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
This publicly available data provides the basis to build a panel of Illinois PU professors 2010-2020; I define a professor as an individual observation by their first plus last name and university pairing, and link this database to IPEDS information on finances for their employing institution.
\begin{table}[h!]
    \onehalfspacing
    \centering
    \caption{IBHED Summary Statistics, Professor Panel 2011--2020.}
    \makebox[\textwidth][c]{\input{tables/illinois-summary.tex}}
    \label{tab:illinois-summary}
\end{table}
The analysis sample focuses on the subset of professors with identified year of hiring between the years 2011 and 2020, representing 1,778 professors in 2011, and 9,099 in 2020.
\autoref{tab:illinois-summary} presents the summary statistics for this sample.\footnote{
    The full data IBHED contain 173,000 observations for every public university professor in Illinois 2010-2020, which representing 16,932 professors in the year 2010 and 15,352 in the year 2020.
    The reasons for restricting to the sample with identified year of hire in the range 2011-2020 is explained in \autoref{sec:iv-model-indiv}.
}

@@

\autoref{fig:funding} exhibits the trends for the mean of US PUs between the years 1990-2020.
We see a rise in total revenues received by PUs (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university. 
Non-institutional revenues refer to revenues coming from external sources, not raised by internal university activities;\footnote{
    US universities raise revenues from many different internal activities, including sales of non-degree related educational courses to publishing deals.
    The non-institutional revenue measure is constructed to consider only revenues resulting from external decisions, including state, local, and federal appropriations and tuition revenues.
    Rates of tuition area considered external since most state governments play a large role in deciding tuition rices for PUs \citep{NBERw23736}.
}
non-institutional revenues rose over this time period almost exclusively thanks to the rise in tuition revenues.
At the same time, total state appropriations stagnated at around \$100 million per year-university for 1990-2008, fell slightly around 2008 and have not recovered.
While PUs experienced a financial stagnation, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2020 with no corresponding decline in any specific component.

@@

At the same time, student enrollment at PUs rose precipitously.
6.2 million students were enrolled in PUs in 1990, and this number rose by 47\% to 9.1 million, with most of the increase occurring over the years 2000-2020.
\begin{figure}[h!]
    \centering
    \caption{Total Student Enrollment, by University Sector, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total.}
        \includegraphics[width=\textwidth]{figures/enrollment-total.png}
        \label{fig:enrollment-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Mean, University Level.}
        \includegraphics[width=\textwidth]{figures/enrollment-mean.png}
        \label{fig:enrollment-mean}
    \end{subfigure}
    \label{fig:enrollment}
\end{figure}

\autoref{fig:enrollment} shows that total enrollment at private universities has also risen over the same time period, but not as drastic in either relative or absolute terms; the mean public university also grew over this time period, from 9,800 students in 1990 to 11,800 in 2020.
This means that per student mean revenues per student (seen in \autoref{fig:mean-funding-fte}) have stagnated across all measures, and particularly fallen from \$11,000 per student in 1990 to less than \$8,000 per student in 2020.
\begin{figure}[!h]
    \centering
    \caption{Total Professor Count per Student, by University Sector, Professor Appointment, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Lecturer}
        \includegraphics[width=\textwidth]{figures/lecturer-fte-perprof.png}
        \label{fig:lecturer-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Assistant.}
        \includegraphics[width=\textwidth]{figures/assistant-fte-perprof.png}
        \label{fig:assistant-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Full.}
        \includegraphics[width=\textwidth]{figures/full-fte-perprof.png}
        \label{fig:full-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{All.}
        \includegraphics[width=\textwidth]{figures/all-fte-perprof.png}
        \label{fig:all-fte-perprof}
    \end{subfigure}
    \label{fig:fte-perprof}
\end{figure}

\autoref{fig:fte-perprof} documents the divergence in faculty composition (per student) between the average private and public university 1990-2020.
Private universities start with  a higher baseline of around 4.5 professors per 100 students, and exhibits yearly variation of less than 0.5 professor per 100 students over the thirty years.
PUs start with 3.9 professors per student, and this number falls to 3.5 primarily in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before thea year 2000, yet this number has fallen by over 20\% in the next 20 years only for PUs: in 2020 the mean PU has 6 fewer full professors per tudent than the mean private university.
At the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990 yet have been utilised by PUs at a higher rate since.

@@

Na\"ively we can analyse the relationship between professor-outcomes $Y_{i,t}$ at university $i$ in year $t$ as a result of university revenues per student $X_{i,t}$.
Professor outcomes include count of professors per student, and average salary, within each university-year.
This model gives $\beta$ the association with these outcomes and university revenues per student,\footnote{
    Note that dividing by student count also implicitly controls for the size of the university, so that this model accounts for yearly variation in professor count and university revenues arising from growth in a university.
}
including fixed effects to control for effects specific to the university and year.
Log$(.)$ transforming the variables improves the interpretability of $\beta$ as an elasticity for professor count per student with respect to institution revenues.
\begin{equation}
    \label{eqn:naivereg}
    Y_{i,t} = \alpha_i + \gamma_t + \beta X_{i,t} + \epsilon_{i,t}
\end{equation}

@@

Estimates for the model where units of analysis are individual professors at Illinois PUs follow.

\begin{table}[!h]
    \onehalfspacing
    \centering
    \caption{2SLS Estimates for Faculty Salaries at Illinois Universities.}
    \makebox[\textwidth][c]{\input{tables/facultysalaries-shock-illinois.tex}}
    \begin{flushleft}
        \footnotesize
        \textbf{Note}: Standard errors are clustered at the institution-year level.
    \end{flushleft}
    \label{tab:facultysalaries-shock-illinois}
\end{table}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
    Cornell University, Department of Economics.}
    \thanks{I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, and Francine Blau for various comments on the project's direction,
    as well as seminar participants at Cornell University (Spring 2022) for helpful
    discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/finances-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.}}
% Title
\title{Stagnating Public University Finances and Faculty Composition}
\date{\today}
\rhead{\today}
\lhead{Stagnating Public University Finances and Faculty Composition}
% Begin
\begin{document}
\clearpage \maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
\noindent
State support for public universities has stagnated the last thirty years, while enrollment has secularly increased.
At the same time, public universities have reduced the number of professors per student on tenure-track or with tenure, while increasing their employment of contingent lecturers.
This analysis relates changes in the faculty composition at public universities to the changes in per student university revenues from state appropriations, addressing endogeneity in state appropriation decisions by a shift-share instrumental variables approach.
An increase of university revenues of 10\%, via a change in state appropriations, increases the number of assistant professors per student by 6.6\% and full professors by 5.3\%, yet decreases the number of lecturers per student by 14.3\%.
Analysis of all the professors hired at Illinois public universities 2011-2021 implies that incumbant professors are not affected by the changes in university revenues, so that changes in faculty composition arises by changes in hiring pratices at public universities.
These results show the compositional effects of stagnating state suport for higher education, and raise questions about the direction that public education is heading thanks to these financial headwinds.

\vspace{0.75cm}
\noindent\textbf{Keywords:}
State and Local Budget and Expenditures,
Higher Education,
Public Sector Labor Markets

\vspace{0.5cm}
\noindent\textbf{JEL Codes:} H72, I23, J45
\end{abstract}
\newpage
\setcounter{page}{1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Starting the paper
\newpage
\doublespacing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
\noindent
Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) the last few decades.
This decline has been shown to lead to worse education and later-life outcomes for students \citep{NBERw23736,NBERw27885}, yet it is not clear what institutional effects or mechanisms driving effects on students.
I show that four-year, degree-granting public universities (PUs) are exposed to tightening financial constraints thanks to falling state financial support (per student), and show this effects the composition of professors at PUs.
Falls in PU revenues, instrumented by state-level finance shocks, lead to a fall in the number of professors per student within a PU, driven by a fall in assistant or tenured professor counts.
US private universities are not exposed to similar financial constraints during this same time period, and do not exhibit any similar changes in the faculty composition, so that stagnating financial support for PUs have implications for the structure of higher education instruction and research in the US.

Universities in the US are widely considered the highest performing in the world,yet there are consequential differences between its universities that operate in the private sector and those established by state governments.
PUs are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average PU received around \$11,600 per enrolled student in 1990, yet only \$8,300 per enrolled student in 2021.\footnote{
    2021 figures are relatively similar to those for year preceding, despite the Covid-19 pnademic and associated financial shocks, since 2021 refers to the academic year 2019-2021 and so the yearly budget was decided in 2019.
    This paper does not directly analyse budget shocks related to the Covid-19 pandemic, or any other events in 2021 and following.
}
This fall is driven by stagnating funding provided by state governments, while total enrollment among PUs rose by 46\% over the same time period.
At the same time, the number of professors per student at private universities stayed relatively stable at 0.045, yet fell from around 0.04 to around 0.35 professors per student at PUs, driven by a fall of over 20\% in the ratio of associate and full (i.e. tenured) professors per student at PUs.\footnote{
    \autoref{fig:fte-perprof} and \autoref{sec:trends} document these trends between private and public universities in full detail.
}

I use a variant of the \cite{NBERw23736,NBERw27885} shift-share instrument for PU revenues from shocks to state appropriations to identify the effect on the faculty composition.
State appropriations for public universities are plausibly endogenous to many outcomes: state governments decide on yearly budgets for their higher education sector, and this process can be influenced by local financial conditions, or even public (and political representatives') perceptions of the state's higher education system.
The shift-share instrument interacts reliance on state funding in a base period with the yearly total appropriations per student in the entire state, to exploit both changes in state support for higher education and how much each university relies on that support.

Findings show that negative changes in university revenues affects faculty composition at public universities, away from tenured and tenure-track professors towards non-tenure track lecturer positions.
A reduction of 10\% in university revenues, via a shock to state appropriations, leads to a fall in 6.6\% in the number of assistant professors per student within a university, 5.3\% for full (tenured) professors, and a reduction in 14.3\% for lecturers.
Additionally, I use individual level data on professors who were hired at Illinois public universities 2011-2021 to investigate whether the shocks to university revenues affected incumbant professors at public universities.
Incumbant professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system, which implies that changes in faculty composition came about by changes in the composition of hiring new professors at the university level.

@@

University professors the core component of this country's educational and research/innovation university system, yet their composition within universities and its relationship with higher education finances has so far not been studied.
\cite{brown2014endowment} presents the closest example by studying how university endowments react to negative financial shocks, exhibiting an association between negative endowment shocks and a fall in tenure-system faculty employed at the university.
\cite{abe2015implications} present a model for university administrator decision-making, and posit that a large component of variation in faculty salaries arises from university politics, and not only economic factors such as outside.
\cite{johnson2009jep,NBERc13879} both empirically document allocation of faculty between research and instruction, and \cite{hemelt2021math} document differences in instructional cost (per student) between departments, primarily thanks to department differences in class size and faculty salaries.
\cite{turner2014impact} document how universities reacted to the Great Recession of 2008, a severe financial shock to state finances, including implementing hiring freezes.
Indeed, Cornell University\footnote{
    Cornell University is, indeed, only partially public, so that this information is not free information, and cannot be presented with aggregate or verified sources of public data on such salaries.
} implemented both hiring freezes and even a nominal salary reduction for professors, in anticipation of oncoming financial shock in early 2021.\footnote{
    The salary cut was not permanent, as the oncoming financial shock turned out to not be as serious as projected, and the salary cut was returned to professors in the year 2021.
}

@@

This analysis finds that faculty composition is meaningfully affected by the last few decades' stagnation in state support for higher education.
Analysis of outcomes for incumbant professors in Illinois implies that the explaining mechanism is changes to hiring by public universities in response to the increased financial constraints.
Previous findings from the literature on the effectiveness of tenured and tenure-track faculty versus contingent faculty in the areas of research and teaching.
Combined with this analysis, these findings provide a plausible mechanism for the deteriorating student ouctomes shown to be a result of stagnating finances at public universities \citep{NBERw23736,NBERw27885}.

@@

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    This statement means that IPEDS does not necessarily cover the universe of US higher education institutions, yet thin practice every PU and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e. 1987 refers to the academic year that ran August 1986 to July 1987.
}
and provide information on the total revenues a university received from every source (including state governments), enrollment number, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990-2017, and manually combine with raw NCES data regarding the year 2018-2021 for all relevant variables.
    Figures for enrollment, faculty counts and salaries come from the raw NCES version of IPEDS for all years to address inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to four-year institutions, as these institutions adhere to a standardised concept of faculty profile -- that is, tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised in four-year, degree-granting institutions.
Additionally, for-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enroll at a four-year institution \citep{mountjoy2022}.\footnote{
    Furthermore, \cite{mountjoy2022} documents the effects of expanding two-year higher education access in the US separately for students who would have and those who would not have otherwise have attended college.
    The high rate of eventual enrollment at four-year institutions among two-year students explains the negative impact of increased access to two-year institutions on students who are diverted away from four-year institutions.
}

@@

\begin{table}[h!]
    \onehalfspacing
    \centering
    \caption{IPEDS Summary Statistics, PUs Panel 1987--2021}
    \makebox[\textwidth][c]{\input{tables/ipeds-summary-fte.tex}}
    \label{tab:ipeds-summary}
\end{table}

IPEDS provides information at the university level, yet lacks information on the distribution of salaries and professor count within a university.\footnote{
    IPEDS provides total paid to salary and total faculty employed per university-year, so that I can investigate average professor salary per university-year with IPEDS data.
    Unfortunately, it seems that this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period \citep{aau2021survey}.
}
To investigate the distribution, I integregate individual-level data for every PU professor in the state of Illinois between the years 2010-2021.
IBHED host the information;
Public Act 96-0266 (effective 1 January 2010) requires that each PU report to the IBHE the base salary and benefits of the president and all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the eight Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
This publicly available data provides the basis to build a panel of Illinois PU professors 2010-2021; I define a professor as an individual observation by their first plus last name and university pairing, and link this database to IPEDS information on finances for their employing institution.
\begin{table}[h!]
    \onehalfspacing
    \centering
    \caption{IBHED Summary Statistics, Professor Panel 2011--2021.}
    \makebox[\textwidth][c]{\input{tables/illinois-summary-rolling.tex}}
    \label{tab:illinois-summary}
\end{table}
The analysis sample focuses on the subset of professors with observed year of hiring between the years 2011 and 2021, representing 1,778 professors in 2011, and 9,099 in 2021.
\autoref{tab:illinois-summary} presents the summary statistics for this sample.\footnote{
    The full data IBHED contain 173,000 observations for every public university professor in Illinois 2010-2021, which representing 16,932 professors in the year 2010 and 15,352 in the year 2021.
    Summary statistics for the entire sample can be see in Appendix \autoref{tab:illinois-summary-entire}.
    The reasons for restricting to the sample with identified year of hire in the range 2011-2021 is explained in \autoref{sec:iv-model-indiv}.
}

@@

\autoref{fig:funding} exhibits the trends for the mean of US PUs between the years 1990-2021.
We see a rise in total revenues received by PUs (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university. 
Non-institutional revenues refer to revenues coming from external sources, not raised by internal university activities;\footnote{
    US universities raise revenues from many different internal activities, including sales of non-degree related educational courses to publishing deals.
    The non-institutional revenue measure is constructed to consider only revenues resulting from external decisions, including state, local, and federal appropriations and tuition revenues.
    Rates of tuition area considered external since most state governments play a large role in deciding tuition rices for PUs \citep{NBERw23736}.
}
non-institutional revenues rose over this time period almost exclusively thanks to the rise in tuition revenues.
At the same time, total state appropriations stagnated at around \$100 million per year-university for 1990-2008, fell slightly around 2008 and have not recovered.
While PUs experienced a financial stagnation, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021 with no corresponding decline in any specific component.

@@

\begin{figure}[h!]
    \centering
    \caption{Total Student Enrollment, by University Sector, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total.}
        \includegraphics[width=\textwidth]{figures/enrollment-total.png}
        \label{fig:enrollment-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Mean, University Level.}
        \includegraphics[width=\textwidth]{figures/enrollment-mean.png}
        \label{fig:enrollment-mean}
    \end{subfigure}
    \label{fig:enrollment}
\end{figure}
At the same time, student enrollment at PUs rose precipitously.
6.2 million students were enrolled in PUs in 1990, and this number rose by 47\% to 9.1 million, with most of the increase occurring over the years 2000-2021.
\autoref{fig:enrollment} shows that total enrollment at private universities has also risen over the same time period, but not as drastic in either relative or absolute terms; the mean public university also grew over this time period, from 9,800 students in 1990 to 11,800 in 2021.
This means that per student mean revenues per student (seen in \autoref{fig:mean-funding-fte}) have stagnated across all measures, and particularly fallen from \$11,000 per student in 1990 to less than \$8,000 per student in 2021.
\begin{figure}[!h]
    \centering
    \caption{Total Professor Count per Student, by University Sector, Professor Appointment, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Lecturer}
        \includegraphics[width=\textwidth]{figures/lecturer-fte-perprof.png}
        \label{fig:lecturer-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Assistant.}
        \includegraphics[width=\textwidth]{figures/assistant-fte-perprof.png}
        \label{fig:assistant-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Full.}
        \includegraphics[width=\textwidth]{figures/full-fte-perprof.png}
        \label{fig:full-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{All.}
        \includegraphics[width=\textwidth]{figures/all-fte-perprof.png}
        \label{fig:all-fte-perprof}
    \end{subfigure}
    \label{fig:fte-perprof}
\end{figure}

\autoref{fig:fte-perprof} documents the divergence in faculty composition (per student) between the average private and public university 1990-2021.
Private universities start with  a higher baseline of around 4.5 professors per 100 students, and exhibits yearly variation of less than 0.5 professor per 100 students over the thirty years.
PUs start with 3.9 professors per student, and this number falls to 3.5 primarily in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for PUs: in 2021 the mean PU has 6 fewer full professors per student than the mean private university.
At the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990 yet have been utilised by PUs at a higher rate since.

\subsection{Trends in Illinois}
\label{sec:trends-illinois}

The state of Illinois particularly exemplifies the stagnation in state support for higher education that the rest of the nation has experienced 1990-2021.
\autoref{fig:illinois-funding} shows the trends for revenues among the seven Illinois public university campuses, including the falling share of state appropriations and substitute towards tuition revenue.

\begin{figure}[h!]
    \centering
    \caption{Mean Revenues among Illinois Public Universities, by Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total, millions \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/illinois-funding-total.png}
        \label{fig:illinois-funding-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Per Enrolled Student, \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/illinois-funding-fte.png}
        \label{fig:illinois-funding-fte}
    \end{subfigure}
    \label{fig:illinois-funding}
\end{figure}

The decade of 2010-2021 has also seen yearly variation in its revenues, particularly around 2016.
In the calendar year 2015 partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting tehniques to continue operation.\footnote{
    Fiscal year 2016 refers to June 205 to June 2016, as is the same for the academic year definition in this analysis. 
    K-12 schools were unaffected thanks to a separate budget agreement allowing non-higher education to operate as before.
}
While state institutions continues to operate, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring revenues to state agencies and universities for the 2018 fiscal year.
This variation in public university revenues, stemming entirely from political disagreements --- not from state decisions regarding higher education and its finances \citep{young2020squandered} --- and via the state appropriations channel, mean that Illinois public universities exhibit sizable changes in their state approriations over 2011-2021, of similar order to those for the rest of the country over 1990-2021.

@@

Na\"ively we can analyse the relationship between professor-outcomes $Y_{i,t}$ at university $i$ in year $t$ as a result of university revenues per student $X_{i,t}$.
The exact measure of university revenues, $X_{it}$ is non-institutional revenues, the sume of university funding from outside sources -- local, federal, and state appropriations plus tuition revenue.\footnote{
    This measure is in line with the \cite{NBERw27885} approach of considering revenues from the state, but does not ignore yearly changes in revenues arising from tuition revenues -- which are shown to be heavily related.
}
Outcomes include count of professors per student, and average salary, within each university-year.
This model gives $\beta$ the association with these outcomes and university revenues per student,\footnote{
    Note that dividing by student count also implicitly controls for the size of the university, so that this model accounts for yearly variation in professor count and university revenues arising from growth in a university.
}
including fixed effects to control for effects specific to the university and year.
Log$(.)$ transforming the variables improves the interpretability of $\beta$ as an elasticity for professor count per student with respect to university revenues.
\begin{equation}
    \label{eqn:naivereg}
    Y_{i,t} = \alpha_i + \gamma_t + \beta X_{i,t} + \epsilon_{i,t}
\end{equation}

@@

The first outcome is total salary for each professor by their position;
\autoref{tab:facultysalaries-shock-illinois} presents 2SLS estimates for the elasticity of total salaries with respect to university revenues.
The results are not distinguishable from zero in any case, which implies that salaries for all years of employment at the university are not causally related to changes in university revenues.
